---
title: "CPDB interaction plots"
output: html_document
---

```{r, include=FALSE}
library(conflicted)

library(ggplot2)
library(ggraph)
library(dplyr)
conflict_prefer("filter", "dplyr")
conflict_prefer("rename", "dplyr")
library(readr)
library(ggrepel)
library(igraph)
```

```{r, include=FALSE}
zeileis_28 = c(
    "#023fa5",
    "#7d87b9",
    "#bec1d4",
    "#d6bcc0",
    "#bb7784",
    "#8e063b",
    "#4a6fe3",
    "#8595e1",
    "#b5bbe3",
    "#e6afb9",
    "#e07b91",
    "#d33f6a",
    "#11c638",
    "#8dd593",
    "#c6dec7",
    "#ead3c6",
    "#f0b98d",
    "#ef9708",
    "#0fcfc0",
    "#9cded6",
    "#d5eae7",
    "#f3e1eb",
    "#f6c4e1",
    "#f79cd4",
    # these last ones were added:
    '#7f7f7f',
    "#c7c7c7",
    "#1CE6FF",
    "#336600"
)

```


```{r, include=FALSE}
cpdb_plot = function(cpdb_res, edge_color="log2_fc", edge_color_scale=scale_edge_color_distiller(palette="RdBu", limits=c(-1,1) * max(abs(conn_meta$log2_fc)))) {
  # make edges df
  edges = cpdb_res |>
    mutate( from=sprintf("%s_%s_ligand", cluster_1, source), to=sprintf("%s_%s_receptor", cluster_2, target)) |>
    select(from, to, name="variable", coef, pvalue, fdr, group, log2_fc) |>
    mutate(log2_fc = if_else(is.infinite(log2_fc) & log2_fc < 0, min(log2_fc[is.finite(log2_fc)]), log2_fc)) |> 
    mutate(log2_fc = if_else(is.infinite(log2_fc) & log2_fc > 0, max(log2_fc[is.finite(log2_fc)]), log2_fc)) |>
    mutate(fdr = if_else(fdr == 0, 1e-255, fdr))
  
  # make nodes df
  nodes_source = cpdb_res |> 
    select(cell_type=cluster_1, gene=source) |>
    mutate(type="ligand") |>
    mutate(name=sprintf("%s_%s_%s", cell_type, gene, type))
  nodes_target = cpdb_res |> 
    select(cell_type=cluster_2, gene=target) |>
    mutate(type="receptor") |>
    mutate(name=sprintf("%s_%s_%s", cell_type, gene, type))
  nodes = bind_rows(nodes_source, nodes_target) |>
    distinct() |>
    select(name, cell_type, gene, type) |>
    filter(name %in% edges$to | name %in% edges$from)

  cell_types = unique(c(cpdb_res$cluster_1, cpdb_res$cluster_2))
  cell_type_lr = unique(paste0(nodes$cell_type, "_", nodes$type))
  
  # construct hierarchy graph for dendrogram
  edges_ct_hierarchy = bind_rows(
    # root
    data.frame(from="root", to=nodes$cell_type),
    nodes |> mutate(to=paste0(cell_type, "_", type)) |> select(from=cell_type, to) |> distinct(),
    nodes |> mutate(from=paste0(cell_type, "_", type), to=name) |> select(from, to)
  )
  nodes_ct_hierarchy = data.frame(name=unique(c(edges_ct_hierarchy$to, edges_ct_hierarchy$from))) |> 
    left_join(nodes, by="name") |> group_by(cell_type) |>  group_modify(function(x, key) { 
      n = nrow(x)
      n1 = floor(n/2)
      n2 = n - 1 - n1
      x$cell_type_label = c(rep(NA, n1), key$cell_type, rep(NA, n2))
      x
    }) |> ungroup() |> select(name, cell_type, gene, type, cell_type_label)
  G_ct =graph_from_data_frame(edges_ct_hierarchy, directed=TRUE, vertices=nodes_ct_hierarchy)


  # info for geom_conn_bundle
  from = match(edges$from, nodes_ct_hierarchy$name)
  to = match(edges$to, nodes_ct_hierarchy$name)
  conn_meta = data.frame(from=nodes_ct_hierarchy[from,]$name, to=nodes_ct_hierarchy[to,]$name) |>
    left_join(edges, by=c("from", "to")) |> 
    mutate(log2_fc = )
  
  # plot using ggraph
  p = ggraph(G_ct, layout="dendrogram", circular=TRUE) + 
    geom_node_point(aes(color=cell_type, shape=type), size=8, stroke=2, fill="white") + 
    geom_conn_bundle(
      data = get_con(from=from, to=to, coef=conn_meta$coef, log2_fc=conn_meta$log2_fc, pvalue=conn_meta$pvalue, group=conn_meta$group, fdr=conn_meta$fdr),
      # aes_string(color=edge_color, width="abs(log2_fc)"),
      aes_string(color=edge_color, width="abs(log10(fdr))"),
      tension=0.4,
      arrow=arrow()) +
    edge_color_scale + 
    scale_edge_width_continuous(range=c(1,5)) + 
    theme_void() + 
    geom_text_repel(aes(x=x, y=y, label=gene), size=5 ) + 
    geom_label_repel(aes(x=x, y=y, nudge_x=1.3*x, nudge_y=1.3*y, label=cell_type_label), xlim=c(-2, 2), size=8, force=3, alpha=.8) + 
    coord_fixed(clip="off") +
    scale_shape_manual(values=c("receptor"=23, "ligand"=19))  +
    scale_color_manual(values=zeileis_28)   +
    theme(legend.position="top")
  p
}

# ggsave("cell2cell.pdf")
```

```{r, include=FALSE}
cpdb_early_advanced = read_tsv("/home/sturm/Downloads/differential_signature_early_advanced_cpdb.tsv")
cpdb_luad_lscc = read_tsv("/home/sturm/Downloads/differential_signature_luad_lscc_cpdb.tsv")
cpdb_infiltration_status = read_tsv("/home/sturm/Downloads/differential_signature_infiltration_status_cpdb.tsv")

# nothing to see here
# cpdb_infiltration_type = read_tsv("/home/sturm/Downloads/differential_signature_infiltration_type_cpdb.tsv")

cpdb_infiltration = read_tsv("/home/sturm/Downloads/differential_signature_patient_immune_infiltration_cpdb.tsv")
cpdb_tumor_normal = read_tsv("/home/sturm/Downloads/differential_signature_tumor_normal_cpdb.tsv")

cpdb_cell_types = read_tsv("/home/sturm/Downloads/differential_signature_cell_types_cpdb.tsv")
```


### Early advanced
```{r, fig.width=22, fig.height=14, message=FALSE, warning=FALSE, results=FALSE}
cpdb_early_advanced |> filter(abs(log2_fc) > 1) |> head(100) |> cpdb_plot()
```

### LUAD LSCC
```{r, fig.width=22, fig.height=14, message=FALSE, warning=FALSE, results=FALSE}
cpdb_luad_lscc |> filter(abs(log2_fc) > 1) |> head(100) |> cpdb_plot()
```

### Infiltration status
```{r, fig.width=22, fig.height=14, message=FALSE, warning=FALSE, results=FALSE}
cpdb_infiltration_status |> filter(abs(log2_fc) > 1) |> head(100) |> cpdb_plot()
```

### Infiltration status (only tumor and structural cells)
```{r, fig.width=22, fig.height=14, message=FALSE, warning=FALSE, results=FALSE}
cpdb_infiltration_status |> filter(abs(log2_fc) > 1, cluster_1 %in% c("Tumor cells", "Stromal", "Endothelial cell")) |> head(100) |> cpdb_plot()
```

### Infiltration type (originating from Tumor, stromal, endothelial cells)
```{r, fig.width=22, fig.height=14, message=FALSE, warning=FALSE, results=FALSE}
cpdb_infiltration |> rename(variable=`...1`) |>
  filter(log2_fc > 1) |> 
  filter(cluster_1 %in% c("Tumor cells", "Stromal", "Endothelial cell"), log2_fc > 1) |>
  mutate(fdr=p.adjust(pvalue, method="fdr")) |> head(100) |>
  cpdb_plot(edge_color="group", edge_color_scale = scale_edge_color_brewer(palette="Set2"))
```

### Tumor vs normal
```{r, fig.width=22, fig.height=14, message=FALSE, warning=FALSE, results=FALSE}
cpdb_tumor_normal |> filter(abs(log2_fc) > 1) |> 
  # filter(cluster_1 %in% c("Tumor cells", "Stromal", "Endothelial cell")) |> 
  head(100) |>
  # clip log fc for better aesthetics
  mutate(log2_fc = sapply(log2_fc, \(x) if_else(x < 0, max(x, -5), min(x, 5)))) |>
  cpdb_plot()
```


## By cell type
```{r, fig.width=22, fig.height=14, message=FALSE, warning=FALSE, results=FALSE}
cpdb_cell_types |> filter(abs(log2_fc) > 1, group %in% c("Stromal", "Endothelial cell", "Tumor cells"), cluster_2 %in% c("Stromal", "Endothelial cell", "Tumor cells")) |> head(150) |> rename(variable=`...1`) |> mutate(cluster_1=group) |> cpdb_plot(edge_color="group", edge_color_scale=scale_edge_color_brewer(palette="Set2"))
```

```{r}
growth_factors = read_csv("../../tables/gene_annotations/genecards_growth_factors.csv", comment = "#")
immune_checkpoints = read_csv("../../tables/gene_annotations/immune_checkpoints.csv", comment="#")
```

## All growth factors (stromal -> tumor)
```{r, fig.width=22, fig.height=14, message=FALSE, warning=FALSE, results=FALSE}
cpdb_cell_types |> rename(variable=`...1`) |> mutate(cluster_1=group) |> filter(group %in% c("Endothelial cell", "Stromal"), cluster_2 == "Tumor cells", source %in% growth_factors$`Gene Symbol`, log2_fc > 2, fdr < 0.01) |> 
  cpdb_plot(edge_color="group", edge_color_scale=scale_edge_color_brewer(palette="Set2"))
```

## All checkpoints (tumor immune)
```{r, fig.width=22, fig.height=14, message=FALSE, warning=FALSE, results=FALSE}
cpdb_cell_types |> rename(variable=`...1`) |> mutate(cluster_1=group) |>
  filter(group %in% c("Tumor cells"), cluster_2 %in% c("T cell CD4", "T cell CD8", "T cell regulatory"), target %in% immune_checkpoints$gene_symbol, log2_fc > 2, fdr < 0.1) |> cpdb_plot()
```

## All checkpoints (LUAD LSCC, tumor immune)
```{r, fig.width=22, fig.height=14, message=FALSE, warning=FALSE, results=FALSE}
cpdb_luad_lscc |> filter(cluster_1 %in% c("Tumor cells"), target %in% immune_checkpoints$gene_symbol) |> 
mutate(fdr=p.adjust(pvalue, method="fdr")) |> filter(fdr < 0.1) |> cpdb_plot()
```

## All checkpoints (infiltration)
```{r, fig.width=22, fig.height=14, message=FALSE, warning=FALSE, results=FALSE}

cpdb_infiltration |> rename(variable=`...1`) |>
  filter(log2_fc > 1) |> 
  filter(cluster_1 %in% c("Tumor cells"), target %in% immune_checkpoints$gene_symbol) |> 
mutate(fdr=p.adjust(pvalue, method="fdr")) |> filter(fdr < 0.1) |>
  cpdb_plot(edge_color="group", edge_color_scale = scale_edge_color_brewer(palette="Set2"))
```


## All checkpoints (early_advanced)
```{r, fig.width=22, fig.height=14, message=FALSE, warning=FALSE, results=FALSE}
cpdb_early_advanced |> filter(cluster_1 %in% c("Tumor cells"), target %in% immune_checkpoints$gene_symbol) |> 
mutate(fdr=p.adjust(pvalue, method="fdr")) |> filter(fdr < 0.1) |> cpdb_plot()
```


